Transportation network with multiple lines

### 2022-04-07 multiple_lines/notes/latex/Notes_trajet.tex
Updated notes
Notes updated

The $(n \times n)$ \emph{origin-destination matrix}, denoted by $\mathbf{N} = (n_{st})$, $n_{st} \geq 0, \; \forall s,t$, contains the flow (e.g. the number of passengers) entering the network in source node $s$ and leaving it in target node $t$. We can denote its margins with 
\begin{align}
\bm{\sigma}_\text{in} := \mathbf{N} \mathbf{e}_n \\
\bm{\sigma}_\text{out} := \mathbf{N}^\top \mathbf{e}_n
\end{align}
where $\mathbf{e}_n$ is the vector of ones of size $n$. The vector $\bm{\sigma}_\text{in} = (\sigma^\text{in}_i)$ is the \emph{vector of flow entering the network} and $\bm{\sigma}_\text{in} = (\sigma^\text{in}_i)$ is the \emph{vector of flow leaving the network}. We have
\begin{equation}
\sigma^\text{in}_{\bullet} = \sigma^\text{out}_{\bullet}.
\end{equation}

@@

Let $\mathcal{P}_{st}$ be the set of \emph{admissible} shortest-paths between $s$ and $t$ on $\mathcal{G}$. We can denote by $\text{P}_{st}(i, j)$ the probability of having edge $(i, j) \in \wp$ when drawing a path $\wp$ from $\mathcal{P}_{st}$. We have 
\begin{equation}
\text{P}_{st}(i, j) := \frac{1}{|\mathcal{P}_{st}|}\sum_{\wp \in \mathcal{P}_{st}} \delta((i, j) \in \wp),
\end{equation}
where $\delta(.)$ designate the indicator function. If we are given an origin-destination matrix $\mathbf{N} = (n_{st})$, we can compute the \emph{shortest-path flow matrix}, noted $\mathbf{X}_\text{sp} = (x^\text{sp}_{ij})$, with:
\begin{equation}
x^\text{sp}_{ij} = \sum_{st} \text{P}_{st}(i, j) n_{st} \label{sp_comp}
\end{equation}
This matrix contains the flow on each edge if we suppose that the flow follows shortest-paths. \\

We can rewrite equation (\ref{sp_comp}) by defining the $(n^2 \times n^2)$ \emph{shortest-path - edge matrix} $\mathbf{S} = (s_{\alpha \beta})$ with
\begin{equation}
s_{\alpha \beta} = \begin{cases}
\text{P}_{st}(i, j) & \text{if } \alpha = s + n(t - 1) \text{ and }  \beta = i + n(j - 1), \\
0 & \text{otherwise}.
\end{cases}
\end{equation}
Then (\ref{sp_comp}) writes
\begin{equation}
\
\end{equation}

@@

We suppose that we know the flow entering and leaving each line, i.e. $\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{out}$ and we are interested in finding trajectories $n_{st}$ of the flow in the network knowing the set $\mathcal{A}$ of admissible pair $(s, t)$, i.e. pairs of nodes where there is a possible use of the network for traveling. This set can be given by the matrix $\mathbf{T} = (t_{st})$ defined by
\begin{equation}
t_{st} = \begin{cases}
1 & \text{if } (s, t) \in \mathcal{A}, \\
0 & \text{otherwise}.
\end{cases}
\end{equation}

\subsection{Algorithm}
Set $\bm{\sigma}^{(0)}_\text{in} = \bm{\rho}_\text{in} + \epsilon$ and $\bm{\sigma}^{(0)}_\text{out} = \bm{\rho}_\text{out} + \epsilon$, with $\epsilon$ a small positive scalar. Until convergence, do:
\begin{enumerate}
	\item Compute $\mathbf{N}^{(i)} = \textbf{Diag}(\mathbf{a}^{(i)}) (\mathbf{T} + \epsilon)\textbf{Diag}(\mathbf{b}^{(i)})$ with iterative fitting, such that $\mathbf{N}^{(i)} \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{in}$ and $\left(\mathbf{N}^{(i)}\right)^\top \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{out}$.
	\item Compute the associated shortest-path flow matrix $\mathbf{X}^{(i)}$ with (\ref{sp_comp}).
	\item Compute $\bm{\sigma}^{(i+1)}_\text{in} = \bm{\rho}_\text{in} -  (\mathbf{X}^{(i)}_\text{B})^\top \mathbf{e}_n$ and $\bm{\sigma}^{(i+1)}_\text{out} = \bm{\rho}_\text{out} - \mathbf{X}^{(i)}_\text{B} \mathbf{e}_n$. 
\end{enumerate}

>>>>

The $(n \times n)$ \emph{origin-destination matrix}, denoted by $\mathbf{N} = (n_{st})$, $n_{st} \geq 0, \; \forall s,t$, contains the flow (e.g. the number of passengers) entering the network in source node $s$ and leaving it in target node $t$. We can denote its margins with 
\begin{align}
\bm{\sigma}_\text{in} := \mathbf{N} \mathbf{e}_n \label{N_rowsum}\\
\bm{\sigma}_\text{out} := \mathbf{N}^\top \mathbf{e}_n \label{N_colsum}
\end{align}
where $\mathbf{e}_n$ is the vector of ones of size $n$. The vector $\bm{\sigma}_\text{in} = (\sigma^\text{in}_i)$ is the \emph{vector of flow entering the network} and $\bm{\sigma}_\text{in} = (\sigma^\text{in}_i)$ is the \emph{vector of flow leaving the network}. We have
\begin{equation}
\sigma^\text{in}_{\bullet} = \sigma^\text{out}_{\bullet}.
\end{equation}

Note that if only $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ are given, a flow matrix $\mathbf{N}$ can be computed relatively to an \emph{orgin-destination affinity matrix} $\mathbf{S} = (s_{st})$, $0 \leq s_{st} \leq 1$, where $s_{st} = 1$ denote a perfect affinity and $s_{st} = 0$ no affinity, through

\begin{equation}
\mathbf{N} = \textbf{Diag}(\mathbf{a}) (\mathbf{S} + \epsilon) \textbf{Diag}(\mathbf{b}),
\end{equation}
where $\textbf{Diag}(.)$ denote the diagonal matrix obtained from a vector, $\epsilon$ a very small quantity, and vectors $\mathbf{a}$ and $\mathbf{b}$ are found through \emph{proportional iterative fitting} algorithm in order to have margin constraints (\ref{N_rowsum}) and (\ref{N_colsum}) respected for $\mathbf{N}$ (a small $\epsilon$ has to be added to $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ if they possess null components).

@@

Let $\mathcal{P}_{st}$ be the set of \emph{admissible} shortest-paths between $s$ and $t$ on $\mathcal{G}$. We can denote by $\text{P}_{st}(i, j)$ the probability of having edge $(i, j) \in \wp$ when drawing a path $\wp$ from $\mathcal{P}_{st}$. We have 
\begin{equation}
\text{P}_{st}(i, j) := \frac{1}{|\mathcal{P}_{st}|}\sum_{\wp \in \mathcal{P}_{st}} \delta((i, j) \in \wp),
\end{equation}
where $\delta(.)$ designate the indicator function. Note that if there is an unique shortest-path between node $s$ and $t$, noted $\wp_{st}$, we have $\text{P}_{st}(i, j) = \delta((i, j) \in \wp_{st})$. \\

If we are given an origin-destination matrix $\mathbf{N} = (n_{st})$, we can compute the \emph{shortest-path flow matrix}, noted $\mathbf{X}_\text{sp} = (x^\text{sp}_{ij})$, with
\begin{equation}
x^\text{sp}_{ij} = \sum_{st} \text{P}_{st}(i, j) n_{st} \label{sp_comp}.
\end{equation}
This matrix contains the flow on each edge if we suppose that the flow follows shortest-paths from origin to destination. \\

We can rewrite equation (\ref{sp_comp}) by defining the $(n^2 \times n^2)$ \emph{shortest-path - edge matrix} $\mathbf{P} = (p_{\alpha \beta})$ with
\begin{equation}
p_{\alpha \beta} = \begin{cases}
\text{P}_{st}(i, j) & \text{if } \alpha = t + n(s - 1) \text{ and }  \beta = j + n(i - 1), \\
0 & \text{otherwise}.
\end{cases}
\end{equation}
Then (\ref{sp_comp}) writes
\begin{equation}
\textbf{vec}(\mathbf{X}_\text{sp}) = \mathbf{P}^\top \textbf{vec}(\mathbf{N}),
\end{equation}
where $\textbf{vec}(.)$ denotes the vectorization function of a matrix, obtained by stacking matrix columns on top of one another. \\

From equation (\ref{sp_comp}), we see that 
\begin{equation}
\frac{\partial x^\text{sp}_{ij}}{\partial n_{st}} = \text{P}_{st}(i, j),
\end{equation}
which equal to $1$ if there is a unique shortest-path between $s$ and $t$ and $(i,j)$ belongs to this path.

@@

\paragraph{The problem:} We suppose that we know the flow entering and leaving each line, i.e. $\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{out}$ and we want to find origin-destination trajectories $n_{st}$. \\

By setting the problem like that, we easily see that it is ill posed. Several solutions exists (e.g. units remain on the same line and follows a first-in/first-out scheme) and we need to add some hypotheses to restrain it.

\paragraph{Hypothesis 1:} Trajectories follow shortest-paths from origin $s$ to destination $t$. \\

\paragraph{Hypothesis 2:} The number of trajectories $n_{st}$ should be as close as possible to $\rho^\text{in}_s \rho^\text{out}_t s_{st}$, where $S = (s_{st})$ is a given affinity matrix between origin and destination nodes. \\

\subsection{Algorithm}
Set $\bm{\sigma}^{(1)}_\text{in} = \bm{\rho}_\text{in}$, $\bm{\sigma}^{(1)}_\text{out} = \bm{\rho}_\text{out}$, and $\mathbf{S}^{(1)} = S$. Until convergence, do:
\begin{enumerate}
	\item Compute $\mathbf{N}^{(i)} = \textbf{Diag}(\mathbf{a}^{(i)}) (\mathbf{S}^{(i)} + \epsilon)\textbf{Diag}(\mathbf{b}^{(i)})$ with iterative fitting, such that $\mathbf{N}^{(i)} \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{in} + \epsilon$ and $\left(\mathbf{N}^{(i)}\right)^\top \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{out}  + \epsilon$.
	\item Compute the associated shortest-path flow matrix $\mathbf{X}^{(i)}$ with (\ref{sp_comp}).
	\item Compute $\bm{\sigma}^{(i+1)}_\text{in} = \bm{\rho}_\text{in} -  (\mathbf{X}^{(i)}_\text{B})^\top \mathbf{e}_n$ and $\bm{\sigma}^{(i+1)}_\text{out} = \bm{\rho}_\text{out} - \mathbf{X}^{(i)}_\text{B} \mathbf{e}_n$. 
	\item Set negative components of $\bm{\sigma}^{(i+1)}_\text{in}$ and $\bm{\sigma}^{(i+1)}_\text{out}$ to 0.
	\item Update $\mathbf{S}^{(i + 1)}$ according to excess flow.
\end{enumerate}
